---
title: "Dataframes - In Class lab"
author: "Damas Kisumbuo"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
```

## Working with built-in data sets

#Load R's "USPersonalExpenditure" dataset using the `data()` function

#This will produce a data frame called `USPersonalExpenditure`

```{r}
data("USPersonalExpenditure")
```


The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
US_Personal_Expenditure <- data.frame(USPersonalExpenditure)
US_Personal_Expenditure
```


What are the current column names of your dataframe?
```{r}
years <- colnames(US_Personal_Expenditure)
years

```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}
Expenditures <- row.names(US_Personal_Expenditure)
Expenditures
```


Add a column "Category" to your data frame that contains the rownames

```{r}
US_Personal_Expenditure$category <- c(Expenditures)
US_Personal_Expenditure
```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}
US_Personal_Expenditure %>% 
  pivot_longer(cols = 2:5,
               names_to = "SECTOR",
               values_to = "YEARS")

```


How much money was spent on personal care in 1940?
```{r}
USPersonalExpenditure["Personal Care", '1940']
```


How much money was spent on Food and Tobacco in 1960?
```{r}
USPersonalExpenditure["Food and Tobacco", "1960"]

```


What was the highest expenditure category in 1960?
```{r}
US_Personal_Expenditure$category[which.max(US_Personal_Expenditure$X1960)]

```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}

lowest_category <- function(year){
  var_name <- paste0("X",year)
  year_of_interest <- US_Personal_Exp[[var_name]]
  min_index <- which.min(year_of_interest)
  US_Personal_Exp$category[]
}



```

```{r}
US_Personal_Expenditure$category[which.min(US_Personal_Expenditure$X1960)]
```


Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}
# sapply(X = US_Personal_Expenditure$category, FUN = lowest_category)
```
